```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=5, fig.path='figure/', echo=FALSE)

#setting system locale, it helps to display plot axis properly and have dates
Sys.setlocale(category = "LC_ALL", locale = "english")

```

---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

I read directly from zip file into dataframe. 
```{r}
df<-read.csv(unz("activity.zip", "activity.csv"), header=T)
```

Variable "date" is transformed from factor type variable to date type. 
```{r}
df$date<-as.Date(df$date, format="%Y-%m-%d")
```

## What is mean total number of steps taken per day?

With help of dplyr package I group data by days and calculate sum of total number 
of steps taken per each day.  
Here's how looks first 10 days.  

```{r include=FALSE}
library(dplyr)
```

```{r group_data_and_calculate}
df_sum_by_date <- group_by(df,date) %>% summarize(steps_per_day=sum(steps))
head(df_sum_by_date,10)
meansteps <- mean(df_sum_by_date$steps_per_day, na.rm=TRUE)
mediansteps <- median(df_sum_by_date$steps_per_day, na.rm=TRUE)
```

```{r histgram_of_data}
hist(df_sum_by_date$steps_per_day, main="Total number of steps taken per day", 
     xlab="Steps per day", breaks=8)
```

Mean total number of steps taken per day is 
**`r format(round(meansteps, 2), nsmall = 2)`**.  
Median of total number of steps taken per day is
**`r format(round(mediansteps, 2), nsmall = 2)`**.

## What is the average daily activity pattern?

We now group data by 5 minutes intervals and calculate average number od steps 
for each interval. Plot of data shows average daily activity.  

```{r average_daily_activity_plot}
#average data, exclude data that is missing
df_avg_by_interval<-group_by(df,interval) %>% 
    summarize(avg_steps=mean(steps, na.rm=TRUE))
library(ggplot2)
ggplot(df_avg_by_interval, aes(x = interval, y = avg_steps)) +
    geom_line(col="blue") + 
    xlab("Time interval by 5 minutes")+
    ylab("Number of average steps")+
    ggtitle("Daily activity")
```

## Imputing missing values

There are missing values in dataset. 
There are **`r sum(is.na(df$steps))`** rows that have "steps" value missing.  

I will populate new dataframe with original data and replace missing values 
with 5-minute average data of days that have value.  

```{r fill_in_missing}
df_new<-df
narows<-is.na(df_new$steps)
df_new[narows,"steps"]<-sapply(df$interval[narows],FUN=function(x){
    df_avg_by_interval$avg_steps[which(df_avg_by_interval$interval==x)]
})
```


```{r calculations_new}
df_sum_by_date_new <- group_by(df_new,date) %>% summarize(steps_per_day=sum(steps))
meansteps_new <- mean(df_sum_by_date_new$steps_per_day, na.rm=TRUE)
mediansteps_new <- median(df_sum_by_date_new$steps_per_day, na.rm=TRUE)
```

Here's histogram of new dataframe with missing values updated.  
```{r histgram_of_new_data}
hist(df_sum_by_date_new$steps_per_day, main="Total number of steps taken per day", 
     xlab="Steps per day", breaks=8)
```

New mean total number of steps taken per day is 
**`r format(round(meansteps_new, 2), nsmall = 2)`**.  
New median of total number of steps taken per day is
**`r format(round(mediansteps_new, 2), nsmall = 2)`**.  

Mean of new data is same as for original data, but median has changed and now is
equal to mean. New data is more centered atound mean.  

## Are there differences in activity patterns between weekdays and weekends?

We want to check for weekday/weekend activity pattern differences. For that purpose
we create new categorical variable in dataset that categorizes date into "weekend/weekday".  

```{r categorical_new}
weekdays=c("Monday","Tuesday","Wednesday","Thursday","Friday")
df_new$weekday<- factor((weekdays(df_new$date) %in% weekdays), 
         levels=c(FALSE, TRUE), labels=c('weekend', 'weekday'))
str(df_new$weekday)
```

Now we group data and compare activities by 5-minute intervals as average steps.  

```{r plot_qlopt_of_activities, fig.height=6}
df_new<-group_by(df_new,interval,weekday) %>% summarise(avg_steps=mean(steps))
qplot(data=df_new, x=interval, y=avg_steps, facets = weekday ~ ., 
      xlab="Time by intervals", ylab="Average steps per interval",
      main="Comparing activities weekday vs weekend", geom="line")
```

